\chapter{Background}

This chapter will investigate music's history as a human activity, and the concepts of participatory and presentational performances will be explained. I will discuss how presentational performance has evolved into its current form and how rock music and technology are bringing participatory elements back into concerts. I will conduct case studies on some specific live music events that are closely related to my interests. Lastly, previous research investigating designing interactive systems for crowds will be reviewed.

\section{Music and Performance}

% Music is a part of being human:
Music is a part of being human. ``The archaeological record," Daniel Levitin (2006) explains, ``shows an uninterrupted record of music making everywhere we find humans, and in every era" (p. 256). Early music making was purely rhythmic, with simple objects being used as percussion instruments. As primitive wind and string instruments were crafted, rhythm was joined by melody. Music making gradually evolved in cultures all over the world, and it has grown to serve many different purposes. In New Guinea it is a gift to one's host; in the Democratic Republic of the Congo it is used to settle lawsuits; Australian aboriginals use music to tell intricate stories; and some African tribes believe repeating musical chants can draw harmful spirits out of inflicted individuals (Jourdain, 1997; Turner, 2011). In modern Western cultures, of course, music is everywhere -- performed at live concerts, scoring film and television, being shared on the Internet. Regardless of culture, there is no denying that music is an inherently social activity.
% Levitin, 2006: Archaeologists have uninterrupted records of music everywhere and every era humans existed
% Jourdain, 1997: Different cultures -- e.g. New Guinea, the Congo, Australian aboriginals -- have different uses for music -- e.g. as a gift, to settle lawsuits, to tell intricate stories
% Turner, 2011: Repetitive, hypnotic mantras are used by African tribes to draw out harmful spirits

% Music as a social activity:
Why did music evolve with humans? Anthropologists believe it may have initially been a tool for social bonding, or perhaps a clever survival method (Jourdain, 1997; Levitin, 2006). A drum circle around a fire could improve a group's coordination, but it may also serve to keep everyone awake and ward off predators. In less primitive societies, music is still valuable as a social activity. Musicologist Thomas Turino (2008) talks about the benefits of making music with others -- what he calls ``sonic bonding." Referencing the work of anthropologist Gregory Bateson, Turino posits that such artistic experiences promote deep connections to others that are crucial for ``social survival." Thus, music making may be a strong tool in forming and developing rewarding relationships. Christopher Small (1998) suggests that live music performance forms communities that represent ideal relationships; perhaps, in the moment, participants forget reality and feel one with those around them. Anthropologist Edith Turner echoes this idea and gives it a name -- \textit{communitas}. ``Communitas," she explains, ``occurs through the readiness of the people -- perhaps from necessity -- to rid themselves of their concern for status and dependence on structures, and see their fellows as they are" (p. 1). Turner identifies communitas at work at sporting events, in the workplace, and even during disasters, but she claims music to be the most reliable source of communitas. Music is ephemeral, emotional, and it cannot be constrained by rules. ``Its life is synonymous with communism, which will spread to all participants and audiences when they get caught up in it" (p. 43). Though we cannot share our bodies with one another, Turner explains, music allows us to share time. It is clear that live music performances can be powerful events; it is important to realize, however, that not all performances are the same.
% Jourdain, 1997: Anthropologists believe music evolved "to strengthen community bonds and resolve conflicts"
% Levitin, 2006:
% * Why did music evolve? Social bonding and coordination, survival methods.
% * A cluster of genes may control both musicality and outgoingness
% Auslander, 1999:
% * Communities are formed based on how the audience interacts, with no dependence on the spectacle at hand
% * Removing the gap between audience and performer stops it from being a performance. Striving for unity only emphasizes the gap. (I disagree.)
% Small, 1998:
% * Community formed at concerts represent ideals, not realities
% * At a live music performance, the values of a certain social group are "explored, affirmed, and celebrated." Meaning comes from the relationships between the people and the venue, among those taking part, and between the sounds being made.
% Turino, 2008:
% * Making music together leads to "sonic bonding"
% * Connecting with others through art is ``crucial for social and ecological survival'' (Bateson)
% * Music and dance promote "flow" -- optimal experience (Csikszentmihalyi)
% * Music making or dancing is a realization of ideal human relationships
% Turner, 2011:
% * Communitas arrises from people ridding themselves of concern for status and structure, readiness to ``see their fellows as they are''
% * The communitas spirit does not take sides, yet it can be ``prostituted'' to produce prejudice
% * A communitas event (here, Carnaval) is not ordinary, it is brief, yet it is vividly remembered
% * ``Music is a fail-safe bearer of communitas''. It is ephemeral, emotional, it breaks the rules. "Its life is synonymous with communism, which will spread to all participants and audiences when they get caught up in it."
% * We cannot share our bodies, but music allows us to share ``precise time''
% * Flow: In control of the situation, yet fully absorbed "so that it seems to be in control of us." A loss of ego. Related to communitas.

% Participatory performances:
In his book \textit{Music as Social Life: The Politics of Participation} (2008), Thomas Turino divides music performance into two categories -- participatory and presentational. Most cultures exhibit some sort of participatory performance. Peruvian communities perform in large groups with each participant either dancing or playing a panpipe or flute; many different religious ceremonies involve singing in unison or in a call-and-response structure; line dancing in North America features choreography that is closely tied to the music. In general, the emphasis is on the intensity of the interactions over the quality of the performance, and participatory performances have characteristics that support this. In a purely participatory performance, Turino explains, ``there are no artist-audience distinctions, only participants and potential participants performing different roles, and the primary goal is to involve the maximum number of people in some performance role" (p. 26). Equality among participants can lower self-consciousness and lead to a more relaxed atmosphere. Having different roles, on the other hand, allows individuals of different skill levels to contribute accordingly. ``Core" and ``elaboration" roles, in Turino's words, cater to less- and more-advanced performers, respectively; core participants keep the performance moving along while elaboration participants add flourish. Another common feature of participatory performances is repetitiveness. This open form allows newcomer participants to easily join in. Additionally, Turino explains, there is a ``security in constancy" that allows performers to become more comfortably immersed in the music. Performances may also incorporate loud volumes, densely overlapped sounds, and loosely tuned instruments as ``cloaking function" to make individuals more comfortable participating. Solos are not common, although sequential soloing sections are sometimes included; karaoke is an example of sequential participatory performance. Overall, participatory performances allow all participants to feel as though they are contributing, and this makes them quite different from presentational performances.
% Turino, 2008:
% * Artists can shift between participatory and presentational performance
% * No artist/audience distinction, only participants possibly performing different roles. Emphasis on sonic and physical interaction. "Heightened social interaction." The "doing" is more important than the end result.
% * Cultural background will determine if a person is comfortable joining a participatory performance
% * Leads to diminished self-consciousness since everyone is equally contributing
% * Different roles of different difficult allow for everyone to feel welcome and achieve flow. "Core" and "elaboration" roles cater to advanced and non-advanced performers respectively.
% * Intensity of participation comes before quality of performance. Inept performers are noticed but not addressed.
% * Participatory performance does not fit well in a capitalist society. It yields social value, not financial value.
% * Open form: Basic motives repeated over and over. Easy for newcomers to join in. "Security in constancy." Can facilitate flow.
% * Hall: Repetition can increase intensity. Synchronicity comforts people. However, Turino believes repetition can "lead to boredom in presentational contexts... Participatory musical styles do not transfer well to presentational stage situations, in spite of nationalists', folklorists', and academics' attempts to bring them into presentational settings." (How do I feel about this?)
% * Wide tuning, loud volumes, and overlapping textures provide a ``cloaking function" that makes people more comfortable participating
% * Virtuosic solos are not common
% * Some participatory performances are sequential -- everyone gets a turn (e.g. Karaoke)
% * Participatory performance may constrain creativity, but it benefits more people overall
% * The doing of activities as a group makes participatory events more engaging than performative

% Presentational performances:
The opposite of the participatory performance is the presentational performance. Here, the performer presents prepared music pieces to an audience that does not directly participate in the performance (Turino, 2008). A typical Western rock concert is a good example: a band on a stage performs rehearsed songs to an attentive audience whose main role is listening to the music. In contrast with the open form of participatory performances, here there is a focus on detail, smoothness, and coherence. These performances are generally closed form; the artist knows how the show will begin and end. While participatory performances rely on constancy, planned contrasts are implemented in presentational performances in order to keep the audience's attention. A rock band will often break up a song with a guitar solo, for instance. Participatory performances foster connections between participants, Turino explains, whereas presentational performances seem to tease a connection between artist and audience without ever realizing it: ``leave them wanting more." Indeed, the goal of a presentational performance is typically to sell as many tickets as possible. It is, in fact, this desire for profit that caused the evolution of presentational performances over the last few centuries.
% Turino, 2008:
% * The performer prepares and presents music to the audience, who does not participate in the performance
% * Closed form. Predetermined beginning, middle, and end.
% * Presentational musicians aim for detail, smoothness, coherence. Predictability and control are key. Some participatory-leaning performers may make adjustments to accommodate the audience.
% * Planned contrasts are implemented in performances to keep the audience interested. This would derail participatory performance.
% * Attractive because it partially fulfills desire to connect to an artist. Corbett suggests concerts are designed to tease but not consummate this connection, "leaving consumers wanting more."


\section{Evolution of the Presentational Performance}

% Presentational performances bred from greed:
Public concerts were virtually unheard of before the 1600s (Jourdain, 1997). Outside of church, commoners rarely had the opportunity to hear ``serious" music, and any other music performance was relaxed and participatory in nature. The `professional' musicians of this era, musicologist Christopher Small (1998) explains, were those hired by aristocrats to accompany them as they played. It was not until the time of the Industrial Revolution that savvy musicians realized that the middle class would see live performances as opportunities to display their newfound wealth. These ``traveling virtuoso-entrepreneurs" made money touring from town to town and performing in local parlours for a fee. By the 1800s, ticketed events were gradually becoming more abundant and ``art music" concerts began transforming the state of the live performance. These were seen as formal events, says Jourdain, and those wealthy enough to attend were expected to follow the established etiquette -- sitting quietly and listening. Music listeners had been turned into consumers.
% Levitin, 2006: Music has only been a spectator activity in the past 500 years
% Jourdain, 1997:
% * Public concerts were virtually unheard of until the 1600s
% * Only after the industrial revolution in the 1800s did common folk gain disposable income and public concerts gain popularity
% * At early operas people talked, ate, and played cards. Even performers would "jabber away" and speak to audience members.
% * Audiences gradually became annoyances to performers. Some conductors would lock out latecomers, silence applause, and ignore requests.
% * Near the end of the 1800s, "art music" performances became more modern -- formal and quiet
% * Some critics call symphony orchestra performances "the epitome of capitalist oppression" -- hierarchical, no audience involvement
% Small, 1998:
% * Musicians were first hired to play with aristocrats. "Traveling virtuoso-entrepreneurs" first started playing public performances when they realized the desire of the middle class to display their wealth.
% * Private performances surfaced around 1730. Ticketed events were not abundant until the 1800s.

% Live music > Recorded music:
Music consumption in the last century has transformed due to a relatively recent development -- recorded music. Wax cylinders, records, cassette tapes, compact discs, and digital files have allowed listeners to experience music performances from the comfort of their homes, and, today, a concert's purpose is typically to `re-present' these recordings. This does not, however, mean that performance has become subordinate to recorded music. Small (1998) maintains that ``performance does not exist in order to present music works, but rather, musical works exist in order to give performers something to perform" (p. 8). Recordings are just blueprints for performance. He also suggests that, since technology now allows people to experience performances at home, the significance of a real live performance actually increases. There are certainly aspects of a live music performance that cannot come from recorded sound. Musicologist Jane Davidson (2006) explains, for instance, that many musicians seem to perform better when in front of an audience; thus, it is possible that the quality of a live performance often trumps that of a recorded performance. Live performances can also be great sources of cultural influence. Ian Inglis (2006), a sociologist studying music performance, agrees: ``In its ability to simultaneously reflect and influence patterns of socio-cultural activity, it is one of the principal avenues along which musical change and innovation can be introduced and recognized" (p. xv). For example, an event as momentary as the 1965 Newport Folk Festival -- where Bob Dylan first went electric -- seemed to change the course of both folk and pop music (Marshall, 2006). The loud guitar startled audience members, and Dylan was booed as he left the stage. Little did that audience know just how plugged in music performances would eventually become.
% Kelly, 2007: Pop performances are typically re-presentations of recorded material ("reviving the musical corpse")
% Jourdain, 1997: Technology put music everywhere, lessening its "nourishment." "We live in an age of widespread musical obesity."
% Auslander, 1999:
% * The desire for live experiences is a product of mediatization. People want to see concerts so they can experience beloved recorded songs in person.
% * Live performances have great cultural value, but Auslander believes mediatized performances will soon be more valuable
% Davidson, 1997: Performers perform better in live situations due to "social facilitation"
% Inglis, 2006: 
% * Live performance is the principal source of musical change/innovation
% * It is unpredictability that makes live better than mediatized performance
% Marshall, 2006: When Bob Dylan went electric at a festival in 1965, it changed the course of pop music
% Small, 1998:
% * Treating the piece of music as the "the supreme reality of art" (Benjamin) implies some fallacies
% * "Musical works exist in order to give performers something to perform"
% * Music is not a thing. "Musicking" is an action.
% * Since people today can hear and see concerts without having to leave their home, the concert "takes on a new and more concentrated ritual significance"

% Technology can enhance presentational performances:
% While speakers are important to amplify the music, screens and visuals have become a key part of concerts...
Technology has been used to enhance presentational performances in many ways. Arena rock concerts make use of arrays of powerful speakers, dense lighting rigs, and multiple giant screens, sometimes also incorporating huge stage pieces and complex mechatronics. The main function of this equipment is to amplify the sights and sounds of the performance. As Kelly (2007) explains, however, it also serves to amplify the persona of the performer. Large screens may show closeups of the performers on stage, but they also often display video clips or abstract visuals designed to reflect the performer's image and communicate underlying themes. Showing clips from Madonna's music videos as she performs, for example, connects the performance to events of the past. This can unify audience members by creating a ``shared past" amongst them, says Kelly. Similar technologies are being used in even more abstract ways by groups like Gorillaz -- an alternative rock band fronted by cartoon characters. Live performances often feature real musicians silhouetted by projections of the fictional band members. German electronic band Kraftwerk has taken it a step farther for live performances of their song ``Robots," leaving the stage entirely and being replaced by their `robotic' analogues for the duration of the song. The fact that audiences will cheer for a stage void of the performers indicates the influence technologies has had on live performance.
% Auslander, 1999: Mediatized performances can offer more intense sensory experiences than traditional performances
% Kelly, 2007:
% * Embodied co-presence of performer and audience is an important convention of pop performances. "Immediacy, intimacy, self-expression and certain kinds of performer-spectator interaction" are expected. The venue becomes "not only a hearing place, but also a seeing place and a being place."
% * "Presence, representation and self-expression" can be either amplified or attenuated by audiovisual technologies in pop shows
% * A 1966 Andy Warhol event (featuring the Velvet Underground and Nico) mixed live performance with projection. Attendees viewed the projections as "part of the show."
% * Imagery framing the performance generally amplifies the performer's presence. This goes against Benjamin's claim that technological intervention will "devalue the here and now of the artwork." Kelly argues that screens showing different perspective in fact amplify the sense of presence. Referring to a Madonna performance, the author states that the juxtaposition in scale between screen and performer actually creates a 'theatrically structured intimacy."
% * Kraftwerk left the stage and were replaced by robots for their song Robots. This "problematises traditional modes of presence in pop-concert reception."
% * Gorillaz is a group fronted by animated characters. They can make a bigger emotional impact than human performers.
% * One Gorillaz performance featured a deceased singer using prerecorded video and audio -- effectively remediating presence. (How can audience presence be remediated?)
% * Displaying clips and themes from her music videos at a Madonna concert creates feelings of a "shared past" in the audience. (How might we create instead a "shared present"?)
% Sexton, 2007:
% * Music is always associated with other media -- performer movement, words in programme notes, album art. Meaning emerges from these relationships.
% * The original phonograph distressed listeners with its lack of visuals. Edison tried to add visual elements to the invention.

% MTV influencing live performance:
So why have visuals become such an important part of rock concerts? As Philip Auslander (1999) explains, live performance and mass media are in competition, and, since mass media is dominating, live performance has responded by imitating its competitor. Live sporting events make use of big screens and instant replays, for instance, and television shows and movies are regularly adapted for the stage. Live music performances, similarly, have been greatly influenced by MTV. As Music Television reached the height of its popularity in the 80s and 90s, concerts began looking more and more like music videos. For example, Madonna's live performance of ``Like A Virgin" at the 1984 MTV Video Music Awards played out like a music video, the singer performing directly to the multiple television cameras and virtually ignoring the large audience beyond the stage (Burns, 2006). Jaap Kooijman (2006) points to Michael Jackson's performance of ``Billie Jean" at the 1983 \textit{Motown 25} concert, where Jackson's outfit and dancing directly referenced the song's music video. The crowd cheered as Jackson exhibited his signature dancing style -- never noticing, or perhaps never caring, that he was lip-synching the vocals throughout. It is clear that technology has been driving the evolution of presentational performance. Modern concerts have truly become spectacles; but should modern concertgoers only be treated like spectators?
% Auslander, 1999:
% * Theatre and mass media (TV, film, recorded music) are rivals. They are not inherently this way, but were made this way over time. Live performance's response to this rivalry has to imitate mass media.
% * "The function of live performance under this new arrangement is to authenticate the video" -- TV > performance (in 1999)
% * The live/mediatized rivalry is not equal. Television is dominant (in 1999). (The Internet is dominant now?)
% Koojiman, 2006:
% * After MTV started in 1981, television became the main promotion tool for music
% * At the Motown 25 performance, Michael Jackson lip-synched "Billie Jean." The focus was on visuals, with Jackson performing the moonwalk for the first time. This was during the rise of MTV, and the performance was essentially a reenactment of the music video.
% Burns, 2006: Madonna's performance at the 1984 MTV Video Music Awards was designed to be a music video, and it literally became one


\section{Return of the Participatory Performance}

% Modern performances are no longer participatory:
The presentational aspects of a rock show are more spectacular than ever. What, then, has happened to the participatory nature in which music traditionally thrives? What is the role of the modern concertgoer? David Horn (2000) provides the following definition:
\begin{quotation}
The popular music event is the sum of a number of smaller occurrences, which might include any or all of the following: the origination or the borrowing of a musical idea; the development of the idea; the conversion or arrangement of the idea into a performable piece; the participation of those (musicians, producers, technicians) whose task is to produce musical sound; the execution or performance of this task; the transmission of the resulting sounds; the hearing of those sounds (p. 28).
\end{quotation}
Out of all of the tasks associated with a live show, the audience is given one -- ``hearing." This spectatorial role developed as performers began ``dominating" over the audience, Jourdain (1997) explains. In the 1800s, conductors began locking latecomers out of the theatre, silencing applause, and ignoring popular requests. Musicians started dressing in matching uniforms, further setting themselves apart from the audience. Audiences may be further dominated by the booming speakers that drown out their voices and the bright lights that make the performer blind to them. Even the audience's primary tool -- applause -- is losing its power. As Baz Kershaw (2001) explains, standing ovations, once rewards that had to be earned by performers, are today dispensed almost without question; the standing ovation is now ``an orgasm of self-congratulation for money so brilliantly spent," he says, rather than a democratic device (p. 144).
% Kelly, 2007: Horn (2000) provides a definition of a pop music performance that belittles the role of the audience
% Jourdain, 1997:
% * Audiences gradually became annoyances to performers. Some conductors would lock out latecomers, silence applause, and ignore requests.
% * Technology has made music less participative. (Is that a challenge?)
% Kelly, 2007: Notions of "identity, authenticity and originality" in pop performance are being challenged. This is allowing for active participation for audience members.
% Small, 1998:
% * Division of talented and non-talented performers is based on a falsehood. We are all capable of creating music.
% * Sharing performance with strangers in the West is abnormal. Most societies' performance feature members of the community.
% * Classical music performance turned audience members into spectators, consumers
% * Audience separated from, "dominated" by performer is a social feature of modern performance. Perhaps a desire to make the perform more "mysterious."
% * 60s/70s folk fests cultivated community, but it was an orchestrated "illusion." Constraints are always necessary.
% * Performers dressing in uniform are separating themselves and their responsibilities
% * Modern orchestra concerts emphasize listening, specialization, formal settings, ticketed admission. They celebrate separation of producer and consumer.
% Turino, 2006:
% * Repetition can "lead to boredom in presentational contexts... Participatory musical styles do not transfer well to presentational stage situations, in spite of nationalists', folklorists', and academics' attempts to bring them into presentational settings." (How do I feel about this?)
% Kershaw, 2001:
% * 20th century London theatre audience have become much more prone to standing ovations than they once were. This may indicate disempowerment and the fall of the theatre as a political environment. Over the 20th century the role of the theatregoer changed from patron to customer. Early audiences would boo or yell approval. After theatre became organized and lavish, a standing ovation became `an orgasm of self-congratulation for money so brilliantly spent... The power of community all but evaporated.`
% * Applause may be an attempt to form a momentary community

% Rock shows encourage participation:
In the face of this established inequality, rock shows are usually environments that encourage audience involvement -- presentational performances with participatory leanings. Crowds can join the music making by clapping or singing along, for example. They can add to the light show by holding up lighters or illuminated cellphones. Perhaps the simplest way for an audience member to become a participant is to move to the music; swaying, dancing, or joining the mosh pit are all methods concertgoers use to connect to performances. These activities may enhance the live experience for concertgoers, but they rarely have a significant impact on the performance as a whole. Some artists, though, go out of their way to allow the audience to affect the course of the performance. Elvis Costello has toured with a ``Wheel of Songs," for instance, inviting audience members to spin the wheel and determine what tune the band will play next. The Flaming Lips dress willing audience members in ridiculous costumes and let them dance on stage for entire concerts. Green Day, playing to thousands of fans, will pull a select few on stage and allow them to play the band's instruments for one song before sending them backstage to enjoy the after party. These are certainly exciting examples, but they are far from involving each and every member of the crowd. We might ask, then -- how can the participatory spirit of the rock performance be amplified?
% Carah, 2010: Iggy Pop was not allowed to invite fans to dance on stage, causing him to vocally object, blaming ``money and TV"
% Some artists are known to take questions from the crowd in between songs. One interviewer mentioned a performer who announced his phone number and responded live to text messages he received.
% There are certainly those who try to draw a line; e.g. Neil Young, Josh Homme. Stage diving has led to serious injury and even death.
% Kelly, 2007:
% * Elvis Costello uses a "Wheel of Songs" that allows spectators to determine the setlist
% * Kraftwerk handed out beeping electronic calculators so the audience could join in during their song Pocket Calculator

% Technology!
As with most things that need amplifying (see the previous section), it seems that technology can provide a solution. Although Jourdain (1997) may consider technology a contributing factor to the decrease in participatory performances, there are recent examples proving that it can do the opposite. Before exploring how technology can help connect artists and fans at live performances, however, it is useful to observe how it is helping connect them elsewhere -- via social media.

% Social media:
Communication studies researcher Nancy Baym (2012) describes the traditional audience-performer relationship as ``parasocial" -- one-sided, with most information flowing from the performer to the audience. The notions of ``rock gods" and ``pop stars" framed musicians as superior beings. Today, however, increased connectivity on the Internet has brought the two parties closer together; some musicians, says Baym, even see their fans more as friends. A multitude of social media services allows performers and audiences to communicate and even collaborate, adding intriguing new dynamics to relationships that had gone unchanged for so long. Artists can directly respond to fan message on Facebook or Twitter, for example, or organize question-and-answer sessions with the Reddit community. Soundcloud allows listeners to post comments as they listen to a song, praising or critiquing certain parts of the track. Many artists invite fans to create remixes of their work, and new services like BLEND.IO facilitate this kind of collaboration. These types of interactions are mutually beneficial; the audience gets to go deeper into the artist's world, and the artist gets to learn more about their listeners.
% Jourdain, 1997: Technology has made music less participative. (Is that a challenge?)
% Baym, 2012:
% * Audience-performer relationships were traditionally `parasocial` -- one-sided
% * Social media has reconfigured this relationship, eliminating `rock gods` and `pop stars`
% * Interviewed musicians feel social media presence is necessary for success
% * Some musicians consider fans their friends that they can count on for support
% * Scalability limits the interactions
% * Distance may maintain ``mystique," but many artists think this is overrated
% * Some fans may be overly emotional, feeling `false intimacy`
% * Social media can, of course, also be a channel for negativity, but this is easy to block
% * Of course, every artist and fan is different
% Listeners are more active; remixes, mashups, fan vids, sharing, commenting, etc. are creating a new paradigm for music enjoyment.
% Karpovich, 2007: Accessible technologies allowed for the popularity of fan vids, non-professional narrative clips inspired by MTV-style music videos

% Tech-enabled concerts:
This connectedness is being increasingly utilized to alter live performances. A simple example is an online event page, where fans can learn more about an upcoming performance, see who else plans on attending, and communicate and share with other fans. Some artists reach out to attendees beforehand as well. For each of their tour dates, alternative rock band Wilco provides an online form where fans can send song requests along with a note or dedication. Experimental performer Rich Aucoin gathers names of concertgoers, promoters, and other artists that are on the bill and projects text on stage individually praising each of them before his show begins. Many music festivals make use of social media channels during performances, displaying messages or photos posted by audience members on Twitter or Instagram. Artists will often share comments or photos from shows as well; Toronto band Born Ruffians, for instance, usually take a photo of their audience from on stage and post it on Facebook, allowing attendees to find and tag their faces in the crowd. The practice of `bootlegging' -- recording concerts to be shared with other fans -- has existed for decades in rock music. While this activity is surrounded by legal issues, many artists openly encourage it; the Grateful Dead and Phish are two well-known examples. Today, some artists, like Bruce Springsteen and the Red Hot Chili Peppers, are saving would-be bootleggers the trouble and providing free professional live recordings online after each show. Video recording is, to some degree, a contemporary version of this issue; fans film some or all of a performance on their personal devices and post the videos online. As with traditional bootleggers, there are musicians who are for and against this practice. It seems to be happening more and more often, for instance, that performers explicitly ask fans to keep their devices in their pockets. Radiohead and The Beastie Boys both embraced the concept, on the other hand, using fan-shot footage to create well-produced concert videos. While a performer's opinions may vary on all of the above-mentioned uses of technology, it is clear that they all allow for new levels of connectedness at a live performance.
% NOTE: Look at your Introduction for material for this paragraph
% The National repost fans' tagged Instagram photos from their shows
% Auslander, 1999: "Rock must constantly change to survive" (Grossberg)
% Wilco allows users to request songs through their website
% Rich Aucoin performs with projects, sometimes incorporating audience members' names. He usually ends shows by throwing a parachute over the crowd and singing/dancing with audience members.
% Born Ruffians take Instagram photos of their audience at every single show, post it to Facebook, and allow everyone to tag themselves
% Sexton, 2007: Interactivity can benefit sound art. Previous interactive sound art has been designed in order to break down barriers between listener and creator. Synchronous and non-synchronous interactive sound art projects exist, each with their positive and negative qualities.
% Bootlegging has seen popularity with certain bands and genres. Today, bands are beginning to offer bootlegs themselves (e.g. RHCP, Springsteen).
% There have been studies done on using technology for new forms of collaborative music creation -- pure participatory performance, where each user is an equal (Gurevich, Tanaka). But this doesn't reflect the nature of artist-fan relationships (?).
% Recording concerts with cellphones is a divisive for concertgoers and artists alike. Some artists enforce cellphone bans, while others like Radiohead and The Beastie Boys have encouraged filming and even made concert films from the collective footage. It is certainly a new way for fans to participate, but does it remove them too much from the experience?
