\chapter{Ethnography}

This chapter describes the ethnographic study that was undertaken. I present and analyze the results of an online questionnaire completed by music fans and three semi-structured interviews conducted with musicians.

\section{Audience Questionnaire}

% Explain the content of the questionnaire:
An online questionnaire was created in order to obtain a sample of modern music fans' opinions on interactive performances. The nine-question survey was completed over one week by ninety-nine participants recruited through links posted on several social media websites. The first few questions informed me of what type of concertgoer each participant was -- asking how often they attend live music performances, which genre they usually see performed, and the size of the venues they frequent. I also asked how often the participants communicate directly with musicians through their social media presences. Next, the questionnaire focused on concert behaviours. Participants were asked in which actions they typically partake at live music performances; choices included applauding, headbanging, and holding up hand-written signs. They were asked how they might like to participate in a performance and what sort of message they would send their favourite performer if they could. I asked for their thoughts on getting involved in performances and bringing new technologies into concert settings.

% Present the direct results:
The results are informative, though not overly surprising. Most respondents attend shows semi-regularly -- one or more per year (46\%) or one or more per month (42\%). Rock music is the most common genre by far, accounting for 42\% of the responses. Venues of small (38\%) and medium (39\%) sizes are more popular than large arenas (22\%). While 47\% of participants stated that they sometimes communicate with artists using social media, 36\% have never done this. The most popular concert actions are applauding (91\%), singing along with the performer (79\%), clapping or stomping to the beat (74\%), and dancing (62\%). The less-common actions include holding up signs (2\%) and booing (2\%). When asked how they might want to participate in a performance in a new way, 47\% said that they would like to choose the songs that are played, while much fewer expressed interest in manipulating visuals and contributing to the music; 27\% of participants stated they did not have interest in directly influencing a live music performance at all. Given the opportunity to communicate with their favourite performer, most participants responded with praise or appreciation (``Thank you," ``I love you!"). Other responses include song requests and suggestions like ``Don't bury the vocals" or ``More rock, less talk!" Most participants (64\%) indicated that they enjoy when performers ask them to participate in a performance -- clapping/singing along or call and response, for example -- while 22\% stated that they are indifferent. Lastly, 59\% said they are excited by the idea of bringing new technologies into a live music setting, with 36\% indicating indifference. The complete set of results can be found in Appendix B.

% Present the results of the in-depth analysis:
Upon deeper analysis of the responses, some correlations were uncovered. There are clear relationships between show-going frequency, typical venue size, and interest in interaction and technology. Participants attending shows more frequently are more likely to visit smaller venues. This group also expressed the most interest in being involved in performances; they are more inclined to interact with their favourite artists via social media; and they are more welcoming to the idea of unfamiliar technology in a concert setting. The opposite, thus, can also be said: participants who go to fewer shows tend to go to larger venues, are more likely to refrain from participating in shows, are less likely to contact artists through social media, and are less interested in new technologies.

\subsection{Analysis}

These results present some possible approaches to addressing the research questions. It is clear, for instance, that most music fans are comfortable moving at live performances; a movement-based input, perhaps similar to Feldmeier and Paradiso's (2007), would not be out of place in this environment. Respondents enjoy clapping and singing along with live music, but, beyond choosing songs which to hear, few are interested in new ways to get involved. Most concertgoers seem satisfied with the current audience-performer relationship. There is, however, a demographic that seems willing to experiment. Music fans who frequent smaller venues are more inclined to participate in performances and more interested in new technologies, and their relatively high social media activity means that they could be considered members of Wikstr\"{o}m's (2013) ``participatory culture." Thus, while the case studies in Chapter 2 involved big-name bands and large audiences, a small rock club may be the best environment for experimenting with participatory technologies.
% Bias: Participants were recruited via social media, so their social media use is skewed

% Questionnaire Outcomes:
% * Small venues seem to attract people that enjoy participating, are interested in technology, etc.
% * Music fans like to move
% * People don't really want a new way to participate


\section{Performer Interviews}

% Explain the recruitment process and describe the participants:
In order to obtain performers' perspectives, three musicians were recruited and separately interviewed: Erik Grice, Blake Enemark, and Christian Hansen. These three were selected because I was familiar with their work and felt that they represented three distinct performance styles within the rock genre. The subjects have different backgrounds as performers and experience playing in different parts of North America for a range of audience sizes. They also have different relationships with modern technology. While their opinions are representative of modern rock musicians, it should be kept in mind that the subjects are themselves music fans; thus, they are also sharing their perspectives as concertgoers.

% Outline the interview questions
After briefly establishing their history as performers, I asked them each about what audience participation means to them. The musicians were shown video of some of my case study subjects (including Xylobands, Wham City Lights, and Kasabian's Interactive Stage Show), and I recorded their reactions and general thoughts on technology-enabled performances. Lastly, the artists were asked if and how they would want to incorporate similar participatory technologies into their own shows. The interviews were semi-structured and therefore quite conversational. Each interview lasted approximately one hour.

\subsection{The Subjects}

\subsubsection{Erik}
Erik Grice grew up outside of Edmonton, Alberta. He performed musical theatre as an adolescent and started playing in bands in his teens. After acting as a vocalist and guitarist in previous bands, Erik now plays drums for Edmonton-based The Fight -- an alternative rock band with soul undertones. The Fight's typical audience size can range from twenty while on tour to over one hundred people at Edmonton shows. Venues are usually small clubs or halls. They have also played acoustic shows at cafes and similar small venues. A typical Fight concert is energetic yet controlled, and audience members can comfortably move to the music.

\subsubsection{Blake}
Blake Enemark is originally from Victoria, British Columbia. He learned to play guitar as a teenager and started performing cover songs in bands. Blake recorded and toured with his alt-country band Forestry throughout 2010. Shortly after, he was recruited to join We Are The City, a Vancouver-based progressive rock band who had just gained nation-wide recognition after winning the \$150 000 Peak Performance award, a revered radio station contest. With this group, Blake played to audiences more sizeable and fanatic than he had ever before, culminating in a performance for around two thousand people in Vancouver's Stanley Park. Blake parted ways with We Are The City after one year. He joined folk group Northcote for a Canadian tour before settling down in Toronto and starting his own project called Snoqualmie. Snoqualmie, described as ``high-gain, sad-sack Canadiana," marks a return to songwriting and more intimate shows for Blake.

\subsubsection{Christian}
Christian Hansen has a theatre background, a graduate of the University of Alberta's acting program. He began playing in bands as a teenager. During his university degree, Christian rediscovered his desire to perform music and began playing acoustic shows around Edmonton. He eventually started performing with his now-wife Molly, incorporating drum machines and prerecorded tracks into his work. This marked the formation of Christian Hansen \& The Autistics. When Christian was forced to perform alone for one show, he was inspired to put all of his energy into his performance; this was a ``lightbulb moment" for him. Christian Hansen \& The Autistics soon gained notoriety around Edmonton for their high-energy shows, and their songs received significant radio play. Christian and Molly moved to Toronto in 2011, shortening their band name to Christian Hansen. They describe their recent music as ``new wave." While they are currently working their way into Toronto's music scene, Christian Hansen typically draw around five hundred excited showgoers when they play in Alberta.

\subsection{Interacting With Audiences}

\subsubsection{Erik}
The Fight encourage some forms of participation at their shows. Their lead singer will often walk out and physically touch audience members, looking them in their eyes as he sings. He will invite the crowd to sing or clap along with certain songs. Erik feels that this sort of participation typically makes for better shows: ``An attentive and participating crowd of fifty people is always going to be better than two hundred people who are standing there with blank faces." He explained that an ideal audience matches the energy that the band exudes. In addition to making the show more enjoyable, he said, this also makes musical flubs less significant to everyone present. Erik feels that it is the band's job to keep the audience's attention. They must make decisions based on the venue, the audience, and the length of the set. Stage banter is usually only implemented to convey pertinent information to the crowd.

The Fight make use of popular social media channels. They use Facebook to advertise, organize contests and giveaways, and share information on their live dates. Twitter is used for communicating with fans and other artists. This has helped them book opening slots for bigger touring bands. At shows, the band hands out cards with links to their social media pages. Erik believes there is a levelling out between artists and fans, and he likes this. He explained that some contemporaries try to maintain personas and seem ``untouchable," but he does not feel that this is beneficial to the artist.

\subsubsection{Blake}
Blake is a self-described introvert. While he is technically the leader of his current band, he expressed a preference for playing in the background. Despite being typically shy with audiences, Blake recognizes the significance of even basic audience participation; singing and clapping along can make one feel like a part of the show. Blake also expressed the impact this might have on performers: ``It would be the most flattering thing in the world for me for someone to sing my song back to me," he said. He noted, however, that different performers react to audience participation differently. A guitarist in his former band, for instance, would not allow audiences to clap during songs. ``There's a fine line," Blake also stated, ``between drunken participation and intentional participation." He acknowledged the prominence of alcohol in live music settings and the importance of alcohol sales at most venues; it is a part of the industry, and it plays a role in how audiences behave.

The internet presence of Blake's current band is ``not very good." Blake recently closed his own Facebook account for personal reasons. Despite this, he acknowledged the importance social media can hold for artists: ``It's bridging gaps that have never been bridged before." Blake had an especially meaningful experience with the Reddit community; after an anonymous user posted a link to his music, he began receiving orders from all over the world -- the southern United States, Poland, Japan. ``You never know who's listening," he said, ``It's empowering and terrifying."

\subsubsection{Christian}
Christian feels that it is the responsibility of the performer alone to ensure a concert is enjoyable. If he puts all of his energy into a show, he explained, all he can hope is that the audience reciprocates: ``If we come in at 200 percent, maybe the audience will get to 100 percent." Christian emphasized the importance of responding to the audience. If they are standing far from the stage, he will encourage them to get closer. If certain individuals are especially invested in the show, he will make eye contact with them and sing directly to them. Christian encourages singalongs and will sometimes hold the microphone in front of audience members who are singing loudly. He sometimes even leaves the stage and makes physical contact with the crowd if they seem particularly comfortable. At their most recent Edmonton concert, Christian entered the audience and performed his last song unplugged, the crowd surrounding him and singing along. For him, this moment was ``amazing, magical." While he acknowledges that every audience is different, Christian feels that getting the audience involved generally increases the intensity of a show.

Christian embraces connecting with fans through social media. He accepts friend requests from fans on his personal Facebook account and does his best to respond to every message he receives. Despite a few negative online experiences, Christian enjoys interacting with fans in this way.

\subsection{Participatory Technologies}

At this point in the interview, the musicians were shown images and videos of projects that were outlined in Chapter 2: PixMob, Xylobands, Wham City Lights, Kasabian's Interactive Stage Show, Plastikman's SYNK, and Amex Unstaged: Usher.

\subsubsection{Erik}
Erik responded negatively to the projects that relied on smartphones; he feels that they would be mostly distracting, and he expressed concern about being responsible for people dropping and damaging their devices. Instead, he favoured the work that had other tangible elements, like the PixMob beach balls. Erik feels that every show should be unique. An attendee should be able to go home and say, ``I was at that one" -- like a souvenir t-shirt with tour dates listed on the back. Some artists post photos and set lists from each of their shows, he explained, and even this makes a show feel personal for those who attended. If the set changes slightly each night, this amplifies the effect, as well as keeping things interesting for the performer. Erik feels that these sorts of technologies should allow people to opt out without affecting the others' experiences. However, he also expressed concern about giving power to all audience members. Groups that hand out percussion instruments like tambourines, for instance, have to deal with participants that do not have rhythm. Erik provided an anecdote about a band that gave miniature harmonicas to audience members, to be played during a particular part of one song; the crowd continued to play the instruments throughout the rest of the set and all of the following bands'. Thus, Erik believes that any audience input would ideally be controlled or edited to limit undesirable content.

\subsubsection{Blake}
Blake commented on the importance of context. An experimental performance may only be successful if the crowd is filled with fans of the artist. He feels that a festival-type setting might be more suitable than a small rock club. Blake remarked on the effectiveness of these projects with large audiences. He recounted a story of seeing U2 perform; the stadium lights were extinguished, and the crowd was instructed to hold up their open cellphones, filling the space with an impressive electric glow. While Blake was concerned that the technologies I showed him may border on gimmicky, he admitted that creating a spectacle has become a significant part of performing: ``You gotta have something that's more than the music," he said, ``You can't be like the Beatles anymore and just record albums and be successful. There has to be an angle." Although Blake admitted the importance of creating a memorable experience for concertgoers, he seemed to lament the current attitude towards live music: ``A lot of people are just there to have a good time. And if you can make them have a good time then you're a good musician. It's a little discouraging." He was also wary of the amount of video recording at modern concerts, explaining that something is lost in a recording. ``A concert's an experience," he said, ``Go and soak it in and remember it and let it sit in your memory."

\subsubsection{Christian}
Christian admires the way the projects all seemed to aim to ``unite" the participants: ``There are not a lot of times when we feel that we're united." He also remarked on the size of the audience in the examples and wondered if similar effects could be replicated with smaller venues and smaller budgets. Christian speculated that perhaps these technological spectacles are especially useful at large shows because the performers are so distant from audience members. Having played a handful of shows at larger venues, he has dealt with open spaces that dissipate energy and guardrails that divide performer and audience; these technologies could be responses to this divide. When asked about incorporating similar technologies in his own shows, Christian was at first hesitant. He expressed concern about giving up the ``rawness" of the performance; new technologies could take audiences out of the show. He was also wary of giving up control of the show. However, he quickly backed up, noting that ``some of the best gigs are when I felt pretty out of control." Giving the audience some control over the lights or even the set list could be appealing, he admitted, although he would not want the crowd dictating the whole structure of the performance. Christian explained that his sets are organized around tension and release; any audience interaction would have to keep the overall flow of the performance in tact. An ideal system would follow a plan while allowing for the spontaneity that makes shows memorable.  Overall, Christian has no reservations about tech-enabled shows. ``It's a natural evolution," he explained; technology has always helped to move rock music forward. ``That's why I love rock and roll ... There are no rules."

\subsection{Analysis}

The subjects all have considerable backgrounds as performers. They have been involved in various types of performances in different venues and for many types of audiences. They all recognized the importance of connecting with fans online and have benefited from doing so. Thus, these are experienced performers who operate within the online participatory culture, making them prime candidates for this study.

A number of themes were present in many of the performers' remarks, and some notable points were brought up. Firstly, all subjects recognized the positive impact that audience involvement can have on the quality of a performance. Erik and Christian specifically mentioned the importance of raising a crowd's energy. Basing a participatory technology on energetic activity could enhance both the audience's and performer's experiences. Each artist also expressed the significance of context. An ideal system would either be designed for a specific type of venue and audience or made adaptive and scalable. It was mentioned that many small venues allow performers to enter the audience and even touch audience members. Perhaps a participatory technology could complement this physical commingling in such venues or emulate it in those where a barrier exists. Blake noted that audience-performer interactions can be especially meaningful for the performer. An effective system should enhance not just the audience's experience but the performer's as well. There were different opinions on the overall purpose of participatory technologies; while Christian focused on their ability to unite a group, Erik was more interested in them serving as unique, lasting souvenirs.  Although a souvenir element seems less relevant to the research questions, it has been implemented before (PixMob, Xylobands) and would likely add value for audience members.

Several concerns were raised by the musicians as well. Blake and Christian expressed worry about new technologies distracting the audience; the spectacle could overshadow the music. A participatory technology should be balanced, complementing the performance without drawing the crowd's (or performer's) attention away from the primary output. A balance should also be struck in how much control the audience is given. An operator could be implemented in a system to monitor and even edit audience inputs. Subjects also suggested that audience members should not be forced to participate if they do not wish to. Thus, an ideal technology would be effective even if only a fraction of the audience is participating. Erik was unexcited by the use of smartphones in a participatory technology, citing them as a possible source of distraction; he also noted that users could be prone to dropping and damaging them. Avoiding the use of expensive personal devices would likely lessen some unease in both audience members and performers. Lastly, the subjects spoke about the influence of alcohol at most rock music performances. A system for audience involvement should account for the possibility of intoxicated participants and incorporate necessary safety precautions.

An obvious yet important outcome of this study was the fact that all musicians are different. The subjects all have substantial experience performing, for instance, but their interactions with audiences differ considerably. While all three musicians were openminded about the projects that they were shown, they undoubtedly have contemporaries that would quickly dismiss such technologies. A participatory technology should reflect the performance style of the artist; ideally, a system would be designed in close cooperation with the performer who would be implementing it.

% Interview Outcomes:
% * A predictable yet notable result of this study was that all musicians are different. The performers interviewed all had substantial experience performing, for instance, but their interactions with audiences differed considerably.
% * All noted that context -- venue size, audience makeup -- is important and can change how decisions are made
% * Performers aim to raise the energy of the audience and feel that this tends to increase the overall quality of the show
% * There was hesitance about giving the audience too much control
% * The projects impressed all of the artists, though certain aspects raised concerns for each. Spectacle could become more important than musical performance.
% * Performers may enter the crowd or physically touch audience members
% * The technologies may be making up for the distance separating performer and audience
% * Audience participation can be perhaps be more meaningful for the performer
% * People should be able opt out of the interaction without affecting others
% * Some performers are distracted by audience involvement
% * Smartphones could be distracting, and people may drop them
% * All subjects recognize the importance of connecting with fans online and have benefited from doing so
% * A participatory technology should reflect that every show is unique
% * Alcohol is an important factor in determining a crowd's behaviour